Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-05-02 overleaf/vectorization.tex
Edits to shorten.

\secref{sec:dependence} describes our dependence analysis and ~\secref{sec:expansion} describes scalar expansion,
which lifts scalars (and arrays) to the corresponding loop dimensionality to create opportunities for vectorization. 
\secref{sec:basic_vectorization} describes our core vectorization algorithm and~\secref{sec:correctness} argues correctness of the vectorization transformation.
\secref{sec:array_writes} extends vectorization with array writes.

@@

\begin{itemize}
\item same-level edge $X\rightarrow Y$ where $X$ and $Y$ are in the same loop nest, say $i,j,k$. E.g., the def-use edge 9 to 10 in the Biometric MPC Source in Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source} is a same-level edge. A same-level edge can be a back-edge in which case a $\phi$ node is the target of the edge.
E.g., 15 to 4 in Biometric is a same-level back-edge.
\item outer-to-inner $X\rightarrow Y$ where $X$ is in an outer loop nest, say $i$, and $Y$ is in an inner one, say $i,j,k$. E.g., 1 to 4 in Biometric forms is an outer-to-inner edge.
%A $\phi$-node can be a source or a target of an outer-to-inner forward edge.
\item inner-to-outer $X\rightarrow Y$ where $X$ is a $phi$-node in an inner loop nest, $i,j,k$, and $Y$ is in the enclosing loop nest $i,j$. E.g., the def-use from 8 to 12 gives rise to an inner-to-outer edge.
%E.g. $\texttt{sum}_0 = \phi(\texttt{sum}_1, 0)$ to $\texttt{c} = \mathit{CMP}(\texttt{sum}_0,\texttt{min}_0)$ is an inner-to-outer edge.
An inner-to-outer edge can be a back-edge as well, in which case both $X$ and $Y$ are $\phi$-nodes with the source $X$ in a loop nested into $Y$'s loop (not necessarily immediately).

@@

\item mixed forward edge $X\rightarrow Y$. $X$ is in some loop $i,j,k$ and $Y$ is in a loop nested into $i,j,k'$. We transform mixed forward edges as follows. Let $\x$ be the variable defined at $X$. We add a variable and assignment $\x' = \x$ immediately after the $i,j,k$ loop. Then we replace the use of $\x$ at $Y$ with $\x'$. This transforms a mixed forward edge into an "inner-to-outer" forward edge followed by an outer-to-inner forward edge. Thus, Basic Vectorization handles one of "same-level", "inner-to-outer", or "outer-to-inner" def-use edges.
 \end{itemize}

@@

An important component of our algorithm is the scalar expansion to the corresponding loop dimensionality, 
which is necessary to expose opportunities for vectorization.
In the Biometric example, {\sf d = S[i*D+j] - C[j]} equiv. to {\sf d = S[i,j] - C[j]}, which gave rise to $N*D$ subtraction operations in the sequential schedule,
is lifted. The argument arrays {\sf S} and {\sf C} are lifted and the scalar {\sf d} is lifted: {\sf d[i,j] = S[i,j] - C[i,j]}.
The algorithm then detects that the statement can be vectorized.

\paragraph{Arrays} Conceptually, we treat all variables as arrays. There are three kinds of arrays. 
%\ana{Arrays is not a good name here.}

@@

\paragraph{Drop dimension}  
The corresponding $\mathit{drop\_dim}$ is carried out when an array written in an inner loop is used in an enclosing loop.
It takes a higher dimensional array, say $i,j,k$ and removes trailing dimensions, say $j,k$:
\[ \mathit{drop\_dim}({\sf A}, (j:J,k:K)) \]
It iterates over $i$ and takes the result at the maximal index of $j$ and $k$, i.e., the result at the last iterations of $j$ and $k$:
\[ {\sf A'}[i] = A[i,J-1,K-1] \]

@@

We present our algorithm followed by an example on Biometric.

\subsubsection{Algorithm}

@@

@@

\WHILE {there are closure $cl_1$ and $cl_2$ that intersect}
\STATE {merge $cl_1$ and $cl_2$}
\ENDWHILE
\FOR {each closure $cl$ (after merge)}
\STATE {create \python{for} $i_d$ \python{in} ... loop}
\STATE {add PHI-nodes in $cl$ to header block}
\STATE {{\color{red}{add target-less PHI-node for A if $cl$ updates array A}}}
\STATE {add statements in $cl$ to loop in some order of dependences}
\STATE \COMMENT { Dimension is not vectorizable: }
\STATE {change ${I_d}$ to $i_d$ in all statements in loop}
\STATE {treat \python{for}-loop as monolith node for def-uses: e.g., some def-use edges become same-level.}
\ENDFOR
{
\color{red}
\FOR {each target-less PHI-node {\sf A!1 = PHI(A!0,A!k)}}
\STATE {in vectorizable stmts, replace use of {\sf A!1} with {\sf A!0}}
\STATE {discard PHI-node if not used in any $cl$, replacing {\sf A!1} with {\sf A!0} or {\sf A!k} appropriately}
\ENDFOR
}
\ENDFOR

>>>>

In this section, we briefly describe the analysis. Details, including edge cases and examples, can be found in the supplementary extended version~\cite{Anon_TR}.
\secref{sec:dependence} describes our dependence analysis and ~\secref{sec:expansion} describes scalar expansion,
which lifts scalars (and arrays) to the corresponding loop dimensionality to create opportunities for vectorization. 
\secref{sec:basic_vectorization} describes our core vectorization algorithm and~\secref{sec:correctness} argues correctness of the vectorization transformation.
\secref{sec:array_writes} extends vectorization with array writes.

@@

\begin{itemize}
\item same-level edge $X\rightarrow Y$ where $X$ and $Y$ are in the same loop nest, say $i,j,k$. E.g., the def-use edge 9 to 10 in the Biometric MPC Source in Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(b) is a same-level edge. 
%A same-level edge can be a back-edge in which case a $\phi$ node is the target of the edge. E.g., 15 to 4 in Biometric is a same-level back-edge.
\item outer-to-inner $X\rightarrow Y$ where $X$ is in an outer loop nest, say $i$, and $Y$ is in an inner one, say $i,j,k$. E.g., 1 to 4 in Biometric forms is an outer-to-inner edge.
%A $\phi$-node can be a source or a target of an outer-to-inner forward edge.
\item inner-to-outer $X\rightarrow Y$ where $X$ is a $phi$-node in an inner loop nest, $i,j,k$, and $Y$ is in the enclosing loop nest $i,j$. E.g., the def-use from 8 to 12 gives rise to an inner-to-outer edge.
%E.g. $\texttt{sum}_0 = \phi(\texttt{sum}_1, 0)$ to $\texttt{c} = \mathit{CMP}(\texttt{sum}_0,\texttt{min}_0)$ is an inner-to-outer edge.
%An inner-to-outer edge can be a back-edge as well, in which case both $X$ and $Y$ are $\phi$-nodes with the source $X$ in a loop nested into $Y$'s loop (not necessarily immediately).

@@

\item mixed forward edge $X\rightarrow Y$. $X$ is in some loop $i,j,k$ and $Y$ is in a loop nested into $i,j,k'$. We transform mixed forward edges as follows. Let $\x$ be the variable defined at $X$. We add a variable and assignment $\x' = \x$ immediately after the $i,j,k$ loop. Then we replace the use of $\x$ at $Y$ with $\x'$. This transforms a mixed forward edge into an "inner-to-outer" forward edge followed by an outer-to-inner forward edge. %Thus, Basic Vectorization handles one of "same-level", "inner-to-outer", or "outer-to-inner" def-use edges.
 \end{itemize}

@@

An important component of our algorithm is expansion of scalars and arrays to the corresponding loop dimensionality, 
which is necessary to expose opportunities for vectorization.
In the Biometric example, {\sf d = S[i*D+j] - C[j]} equiv. to {\sf d = S[i,j] - C[j]}, which gave rise to $N*D$ subtraction operations in the sequential schedule,
is lifted. The argument arrays {\sf S} and {\sf C} are lifted and the scalar {\sf d} is lifted: {\sf d[i,j] = S[i,j] - C[i,j]}.
The algorithm then detects that the statement can be vectorized.

Expansion (and also reduction) is done with the $\mathit{raise\_dim}$ and $\mathit{drop\_dim}$ functions, 
which we believe are standard. $\mathit{raise\_dim}$ takes the original array, the access function 
$f(i,j,k)$ in loop nest $i,j,k$ and the loop bounds $((i\!:\!I),(j\!:\!J),(k\!:\!K))$ and produces a new 3-dimensional array 
${\sf A'}$ by iterating over $i,j,k$ and setting each element of  ${\sf A'}$:
\[ \mathit{raise\_dim}(A, f(i,j,k), ((i\!:\!I),(j\!:\!J),(k\!:\!K))): {\sf A'}[i,j,k] =  {\sf A}[f(i,j,k)] \]
Raise dimension applies when reshaping input arrays and also, at outer-to-inner def-use edges.
Analogously $\mathit{drop\_dim}$ applies at inner-to-outer def-use edges.
It takes a higher dimensional array, say $i,j,k$ and removes trailing dimensions, say $j,k$. 
It iterates over $i$ and takes the result at the maximal index of $j$ and $k$, i.e., the result at the last iterations of $j$ and $k$:
\[ \mathit{drop\_dim}({\sf A}, (j\!:\!J,k\!:\!K)): {\sf A'}[i] = A[i,J-1,K-1] \] 
%It iterates over $i$ and takes the result at the maximal index of $j$ and $k$, i.e., the result at the last iterations of $j$ and $k$:
%\[ {\sf A'}[i] = A[i,J-1,K-1] \]

\begin{comment}
\paragraph{Arrays} Conceptually, we treat all variables as arrays. There are three kinds of arrays. 
%\ana{Arrays is not a good name here.}

@@

\paragraph{Drop dimension}  
The corresponding $\mathit{drop\_dim}$ is carried out when an array written in an inner loop is used in an enclosing loop.
It takes a higher dimensional array, say $i,j,k$ and removes trailing dimensions, say $j,k$:
\[ \mathit{drop\_dim}({\sf A}, (j:J,k:K)) \]
It iterates over $i$ and takes the result at the maximal index of $j$ and $k$, i.e., the result at the last iterations of $j$ and $k$:
\[ {\sf A'}[i] = A[i,J-1,K-1] \]
\end{comment}

@@

%\subsubsection{Algorithm}

@@

\paragraph{Pseudocode:}

@@

\WHILE {there are closure $cl_1$ and $cl_2$ that intersect}
\STATE {merge $cl_1$ and $cl_2$}
\ENDWHILE
\FOR {each closure $cl$ (after merge)}
\STATE {create \python{for} $i_d$ \python{in} ... loop} \COMMENT{ i.e., MOTION loop }
\STATE {add PHI-nodes in $cl$ to header block} 
\STATE {{\color{red}{add target-less PHI-node for A if $cl$ updates array A}}}
\STATE {add statements in $cl$ to loop in some order of dependences}
\STATE \COMMENT { Dimension is not vectorizable: }
\STATE {change ${I_d}$ to $i_d$ in all statements in loop}
\STATE {treat \python{for}-loop as monolith node for def-uses: e.g., some def-use edges become same-level.}
\ENDFOR
{
\color{red}
\FOR {each target-less PHI-node {\sf A!1 = PHI(A!0,A!k)}}
\STATE {in vectorizable stmts, replace use of {\sf A!1} with {\sf A!0}}
\STATE {discard PHI-node if not used in any $cl$, replacing {\sf A!1} with {\sf A!0} or {\sf A!k} appropriately}
\ENDFOR
}
\ENDFOR
